Non-word repetition in Yélî Dnye

### 2021-01-29 manuscript.Rmd
new version for Middy

@@

- read over whole thing -- does the logic sound ok?
- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four aspects of our understanding of phonological development.

First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR. Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely language. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development. Our reading of previous NWR research is that there are variable effects of length between populations. For instance, @jabere2018xperiment reports an average of ~96% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but ~81% for items 3 syllables long. In contrast, @piazzalunga2019articulatory observe no decline in performance in similarly-aged Italian learners, with a score of 84% for 2 syllables versus 85% for 3 syllables. It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To attempt to see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  @jabere2018xperiment; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

pl<- read.csv("previous_lit.csv")

@@

@@

jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)
plot(pl_stacked$Median~jit_age,col=mycols[pl_stacked$Language],pch=as.character(pl_stacked$Length),
     xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100))

for(thislang in levels(pl_stacked$Language)) for(thislength in levels(factor(pl_stacked$Length[pl_stacked$Language== thislang]))){
  myx=pl_stacked$Age[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
  myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength),lwd=mywd[thislang]) #join points of same length & lang
  mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myhb=myy+pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  for(i in 1:length(mylb)) {
    mylty=as.numeric(as.character(thislength))
    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
    }
}

pl$dif<-pl$X2-pl$X3
difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)

@@

Our reading of the previous literature is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only `r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese `r round(difs[difs$Group.1=="Cantonese","x"])`% versus English `r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.

Third, there are ongoing discussions as to what the key factors structuring individual variation are. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], and/or outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge]. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned. Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English. And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF. **return to this after reading lit**

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference.

@@

dim(mat_pc2)

@@

To address whether errors were better predicted by cross-linguistic or within-language frequency, we calculated a proportion of productions that were correct for each phone (regardless of the type of error or the substitution pattern). Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other. Although a direct test is missing, the correlation with cross-linguistic frequency [r(`r phone_cor_xl$parameter`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`]  was greater than that with within-language frequency [r(`r phone_cor_wl$parameter`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`].

@@

We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments). We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one that has received less attention, regarding the possible relationship between phone frequency and NWR scores. An additional overarching goal was to revisit this task, which is very commonly used to document phonological development, particularly in children raised in urban settings, and who are learning IndoEuropean languages. We will thus consider our results not only in light of previous work, but also with attention to potential linguistic and population differences.

@@

**FIX LOGIC IN THIS PARA** Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks. Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores. Typologically common sounds are associated with higher performance on a handful of other tasks (**REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?**) though to our knowledge this has not yet been tested with non-word repetition. Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts. We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others. Indeed, we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often. This effect is large, with a magnitude more than twice the size of the effect of participant age. This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies. In sum, typological frequency effects, which have been found in other measurements of phonological processing. appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.

**REWRITE THIS PARA** ~With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs. oral vowels and simple vs. complex consonants, revealed no remarkable bias in error types.~ That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.

@@

We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item. Based on previous work, we had predicted that children would have higher NWR scores for shorter items. That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words. Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **M2A: Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.""). I don't have a reference for this, can you please finish the thought or nix this bit?**. Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long. While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:fig-prevlit)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant. In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length. First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon. Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors. **ADD WHAT LANGUAGES WOULD BE IDEAL TO TEST THIS HYPOTHESIS**

@@

A review of previous work (see Introduction) suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR. We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta-analytic efforts aggregating over smaller studies such as ours.

Following prior work, we expected that NWR scores would increase with participant age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or participant gender [@chiat2007preschool] on NWR scores. In addition to this prior work, education on Rossel Island, while generally highly valued, is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation. There is also limited variation in maternal education across the families in the region of the island where we sampled. We therefore expected little evidence for impact of either participant gender or maternal education in the present study. On the other hand, these predictors have established effects on other language development measures (**REFS: M2A: Alex go ahead and pick your faves here**). So to the extent that NWR scores share causal links to gender-based differences in development and maternal linguistic input with these other language outcome measures, we might then expect these factors to appear in NWR data. In fact, participant gender and maternal education correlated with NWR score at about r~.1, which is small.

Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect], presumably because later-born children receive a smaller share of maternal input than their older siblings. Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers. Therefore we expected that any effects of birth order on NWR would be attenuated in this context. In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score. However, the effect size was larger than that found for the other factors, at r~.2, and thus we believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.

>>>>

r2d<-function(r){(2*abs(r))/sqrt(1-abs(r)^2)}

d2r=function(d,n1,n2) {
  a=(n1+n2)^2/(n1*n2)
  d/sqrt(d^2 +a)}
```

@@

- read over whole thing -- does the logic sound ok?
- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
- do you mind if I change all the figures to say NWR scores rather than NWR Scores? In the text, we use lower case.

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas.

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely language. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  @jabere2018xperiment; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

```{r prep-prevlit, echo=F}
pl<- read.csv("previous_lit.csv")

@@

pl$dif<-pl$X2-pl$X3
difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
```

@@

#just a trick to jitter age
jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)

plot(1,xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))

for(thislang in levels(pl_stacked$Language)) for(thislength in levels(factor(pl_stacked$Length[pl_stacked$Language== thislang]))){
  myx=pl_stacked$Age[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
  points(pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]~myx,col=mycols[thislang],pch=as.character(thislength))
  myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength),lwd=mywd[thislang]) #join points of same length & lang
  mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myhb=myy+pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  for(i in 1:length(mylb)) {
    mylty=as.numeric(as.character(thislength))
    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
    }
}

@@

Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge]. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned. Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory which lead any error to result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.

@@

#dim(mat_pc2)

@@

To address whether errors were better predicted by cross-linguistic or within-language frequency, we calculated a proportion of productions that were correct for each phone (regardless of the type of error or the substitution pattern). Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other. Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency [r(`r phone_cor_xl$statistic`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`]  was greater than that with within-language frequency [r(`r phone_cor_wl$statistic`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`].

@@

We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments). We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores. An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and other that will be discussed below). We consider implications of our results on each of these four research areas in turn.

@@

Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks. Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores. It would stand to reason that typologically common sounds are associated with higher performance, but to our knowledge this has not yet been tested with NWR. There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts. One could then predict that this effect should be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.

And yet, we found a robust effect of average segmental cross-linguistic frequency on NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often. This effect was large, with a magnitude more than twice the size of the effect of participant age. Moreover, this significant effect remained even once also accounting for the frequencies of these segments in Yélî Dnye children's input. An analysis of the substitutions made by children also aligned with this interpretation, with more common sounds being substituted for less common ones.

We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development. We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this effect is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR. We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the fine-grained phonetic representations in the perception-production loop [as in e.g., @edwards2004interaction].

@@

We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item. In broad terms, children should have higher NWR scores for shorter items. That said, previous work summarized in the Introduction has shown both very small [e.g., @piazzalunga2019articulatory] and very large [e.g., @cristia2020infant] effects of stimulus length. Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long. The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant, and is small and inconsistent across ages (see Figure \@ref(fig:fig-prevlit)). We do not have a good explanation for why samples in the literature vary so much in terms of the size of length effects, but two possibilities are that this is not truly a length effect but a confound with some other aspect of the stimuli, or that there is variation in phonological representations that is poorly understood. We explain each idea in turn.

First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli. For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants -- which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words. Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower. If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.

Second, NWR is often described as a task that tests flexible perception-production, and as such it is unclear why length effects should be observed at all. However, it is possible that NWR relies on more specific aspects of perception-production, in ways that are dependent on stimulus length. A hint in this direction comes from work on illiterate adults, who can be extremely accurate when repeating short non-words, but whose NWR scores are markedly lower for longer items. In a longitudinal study on Portuguese-speaking adults who were learning to read, @kolinsky2018completely found that, before reading training, the group scored 12.5% on 5-syllable items, whereas after 3 months of training, they scored 62.5% on such long items, whereas performance was at 100% for monosyllables throughout. Given that as adults they had fully acquired their native language, and obviously they had flexible perception-production schemes that allowed them to repeat new monosyllables perfectly, the change that occurred in those three months must relate to something else in their phonological skills, something that is not essential to speak a language natively. Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills. Since as stated this hypothesis is under-specified, further conceptual and empirical work are both needed.

@@

Our review of previous work in the Introduction suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR. We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta- or mega-analytic efforts aggregating over studies.

In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but note  @cristia2020infant is an exception]. Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores. We did not expect large effects of maternal education in our sample because of two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. Similarly, NWR scores may not vary greatly with participant gender according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.

Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings. Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers. Therefore we expected that any effects of birth order on NWR would be attenuated in this context. In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score. However, the effect size was larger than that found for the other two factors and it is far from negligible, at r~.2 or Cohen's d~`r r2d(.2)`. In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the effects we found are larger. We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
